---
title: "Data Wrangling and Analysis in R"
output:
  html_document:
    df_print: paged
---


## 1. Installing packages 
In R, packages are a collection of functions that provide functinalities not given by the R core functionalities. We'll install tidyverse packages which will provide us with packages for performing data manipulation and analysis.

#### Let's install tidyverse packages in Rstudio
```{r}
install.packages("tidyverse")
install.packages("readr")
```

#### Load tidyverse for use in our notebook
```{r}
library(tidyverse)
library(readr)
```

## 2. Load dataset
MTN Telecom offers mobile and internet services to its customers. These services include phone, multiple lines, internet, online security, online backup, device protection, tech support, and streaming TV and movies.

You have been provided with the current customer data. Since you will be working towards a descriptive report than a predictive one, you decide to think critically of the kind of questions that would help you craft customer retention programs. You then later use the given data set to answer your questions but before you start, you reading, explore, clean and visualise your dataset.


```{r}
# Import, Load dataset and preview its summarized information.

MTN_df <- read.csv("https://bit.ly/2ZlpzjF")
head(MTN_df)


```

```{r}
#check the no of rows and cols
dim(MTN_df)
```

## 3. Data cleaning

### Data understanding
```{r}
# Step1: get column names and perform standardisation
names(MTN_df)

```

```{r}
#Step 2: convert all column names to lower case
names(MTN_df) <- tolower(names(MTN_df))
names(MTN_df)

```

```{r}
#Step3: Remove all leading and trailing spaces
names(MTN_df) <- trimws(names(MTN_df), which = c("both"))
names(MTN_df)
```

### Removing Irrelevant data
```{r}
#check for duplicates
MTN_df[duplicated(MTN_df), ]
```

```{r}
#remove duplicates from the dataframe
MTN_df <- unique(MTN_df)

#confirm all duplicates are removed
MTN_df[duplicated(MTN_df), ]
```
### handle missing data
```{r}
#check if there is missing data from the df
colSums(is.na(MTN_df))
```

```{r}
#Remove all records with missing data
clean_df <- na.omit(MTN_df)

#confirm no missing values
colSums(is.na(clean_df))

unique(clean_df$seniorcitizen)


```

```{r}
# column senior_citizen should be a boolean
clean_df$seniorcitizen[clean_df$seniorcitizen == 0] <- 'No'
clean_df$seniorcitizen[clean_df$seniorcitizen == 1] <- 'Yes'

#Check unique values
unique(clean_df$seniorcitizen)
```

```{r}
#resolving the issues in the 'paymentmethod' variable

# - Replace 'Mailed checkkk' with 'Mailed check',
clean_df$paymentmethod[clean_df$paymentmethod == "Mailed checkkk"] <- "Mailed check"

# - Replace 'Electronic checkk', 'Electronic check'
clean_df$paymentmethod[clean_df$paymentmethod == "Electronic checkk"] <- "Electronic check"

#confirm all replacements have been done
unique(clean_df$paymentmethod)
```

```{r}
# visual distribution of the outliers using a box plot

#prepare data
tenure <- clean_df$tenure
monthly_charge <- clean_df$monthlycharges
Total_charge <- clean_df$totalcharges


#plot the chart
boxplot(tenure,
        names = "tenure",
        xlab = "Customers",
        ylab = "Tenure", 
        main = "boxplot with customer tenure")

boxplot(monthly_charge,
        names = "monthly",
        xlab = "Customers",
        ylab = "Billed Amount", 
        main = "boxplot with monthly charging data")

boxplot(Total_charge,
        names = "Total",
        xlab = "Customers",
        ylab = "Billed Amount", 
        main = "boxplot with Total charging data")

```

### Dealing with outliers

```{r}
#Removing the outliers using IQR method
dim(clean_df)
#calculate lower and upper percentiles
lower_bound <- quantile(clean_df$tenure, 0.025)
upper_bound <- quantile(clean_df$tenure, 0.975)
lower_bound
upper_bound

#getting values that are not outliers
tenure_not_outlier_values <- which(clean_df$tenure > lower_bound & clean_df$tenure < upper_bound)

IQR_df<- clean_df[tenure_not_outlier_values,]

dim(IQR_df)

```

```{r}
lower_bound <- quantile(IQR_df$monthlycharges, 0.025)
upper_bound <- quantile(IQR_df$monthlycharges, 0.975)
lower_bound
upper_bound

#getting values that are not outliers
not_outlier_values <- which(IQR_df$monthlycharges > lower_bound & IQR_df$monthlycharges < upper_bound)

IQR2_df<- IQR_df[not_outlier_values, ]

#check values after removing outliers
dim(IQR2_df)
```

```{r}
lower_bound <- quantile(IQR2_df$totalcharges, 0.025)
upper_bound <- quantile(IQR2_df$totalcharges, 0.975)
lower_bound
upper_bound

#getting values that are not outliers
not_outlier_values <- which(IQR2_df$totalcharges > lower_bound & IQR2_df$monthlycharges < upper_bound)

Telecom_clean_df<- IQR2_df[not_outlier_values, ]

#check the new df
dim(Telecom_clean_df)
```


## 4. Data analysis
Since our research question requires us to focus on how we can retain customers, the main variable to work with is the "churn" variable.

### Research specific analysis
What percentage of customers from our dataset churned?

```{r}
# plotting a pie chart to investigate whether MTN customers churned

# prepare data
df <- Telecom_clean_df %>% 
  group_by(churn) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

# df_1 <- Telecom_clean_df %>%
#   count(churn)
# df_1

#Plot
ggplot(df, aes(x = "", y = perc, fill = churn)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Churn", 
    title = "Churn Distribution"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```



```{r}
# we investigate the distribution of churn by senior citizen
# and recording our observations.
df_snr <- Telecom_clean_df %>% 
  filter(seniorcitizen =='Yes') %>%
  group_by(churn) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_snr

#plot the pie chart
ggplot(df_snr, aes(x = "", y = perc, fill = churn)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Churn", 
    title = "Churn Distribution among senior citizens"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```

```{r}
# plotting the distribution of churn by partner
# and recording our observations.

df_ptr <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(partner) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_ptr

#plot
ggplot(df_ptr, aes(x = "", y = perc, fill = partner)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Partner", 
    title = "Churn Distribution by partner"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```

```{r}
# plotting the distribution of churn by dependents 
# and then record your observations
df_dpt <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(dependents) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_dpt

#plot
ggplot(df_dpt, aes(x = "", y = perc, fill = dependents)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Dependents", 
    title = "Churn Distribution by Dependents"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```

```{r}
# plotting the distribution of churn by phone service 
# then recording your observations.

df_phn <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(phoneservice) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_phn

#plot
ggplot(df_phn, aes(x = "", y = perc, fill = phoneservice)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Phone Service", 
    title = "Churn Distribution by Phone Service "
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by internet service 
# and later record our observations.

df_int <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(internetservice) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_int

#plot
ggplot(df_int, aes(x = "", y = perc, fill = internetservice)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Churn", 
    title = "Churn Distribution by Internet Service"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by online security and then record
# our observations below:
df_sec <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(onlinesecurity) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_sec

#plot
ggplot(df_sec, aes(x = "", y = perc, fill = onlinesecurity)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Online Security", 
    title = "Churn Distribution by Online Security"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```


```{r}
# plotting the distribution of churn by online backup
df_bckup <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(onlinebackup) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_bckup

#plot
ggplot(df_bckup, aes(x = "", y = perc, fill = onlinebackup)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Online Backup", 
    title = "Churn Distribution by Onlinebackup"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by tech support
df_techspt <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(techsupport) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_techspt

#plot
ggplot(df_techspt, aes(x = "", y = perc, fill = techsupport)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Tech Support", 
    title = "Churn Distribution by Tech Support"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by streaming tv

df_tv <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(streamingtv) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_tv

#plot
ggplot(df_tv, aes(x = "", y = perc, fill = streamingtv)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Streaming TV", 
    title = "Churn Distribution by Streaming TV"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```

```{r}
# plotting the distribution of churn by streaming movies
df_movies <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(streamingmovies) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_movies

#plot
ggplot(df_movies, aes(x = "", y = perc, fill = streamingmovies)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Streaming Movies", 
    title = "Churn Distribution by Streaming Movies"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by contract
df_ctrct <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(contract) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_ctrct

#plot
ggplot(df_ctrct, aes(x = "", y = perc, fill = contract)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Contract", 
    title = "Churn Distribution by Contract"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```


```{r}
# plotting the distribution of churn by paperless billing
df_ppr <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(paperlessbilling) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_ppr

#plot
ggplot(df_ppr, aes(x = "", y = perc, fill = paperlessbilling)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Paperless Billing", 
    title = "Churn Distribution by Paperless billing"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()
```

```{r}
# plotting the distribution of churn by payment method

df_pay <- Telecom_clean_df %>% 
  filter(churn =='Yes') %>%
  group_by(paymentmethod) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
df_pay

#plot
ggplot(df_pay, aes(x = "", y = perc, fill = paymentmethod)) +
  geom_col() +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  labs(
    fill = "Payment Method", 
    title = "Churn Distribution by Payment Method"
  ) +
  coord_polar(theta = "y")+ 
  theme_void()

```

##  Observations

We observe the following:
  Electronic check payment method had the highest churn rate at 57.9% while mailed check had the least churn rate at 11.2%
  78% of paperless billing customers churned



```{r}
# creating a data frame for gender
# previewing our data set's gender distribution

# Create the summary table
gender_df <- Telecom_clean_df %>% 
  count(gender)
gender_df

# Plotting the chart
ggplot(gender_df, aes(x = gender, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#006756"
  ) +
  labs(
    title = "Gender distribution for MTN customer base", 
    x = "Gender", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# plotting distribution of senior citizens

# Create the summary table
senior_df <- Telecom_clean_df %>% 
  count(seniorcitizen)
senior_df

# Plotting the chart
ggplot(senior_df, aes(x = seniorcitizen, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Senior Citizen distribution for MTN customer base", 
    x = "Senior Citizen", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# Plotting a distribution of subscribers with partners and those without

# Create the summary table
partner_df <- Telecom_clean_df %>% 
  count(partner)
partner_df

# Plotting the chart
ggplot(partner_df, aes(x = partner, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#009919"
  ) +
  labs(
    title = "Partner distribution for MTN customer base", 
    x = "Partner", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# plotting a distribution of customers with dependents and those without

# Create the summary table
dependents_df <- Telecom_clean_df %>% 
  count(dependents)
dependents_df

# Plotting the chart
ggplot(dependents_df, aes(x = dependents, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#828909"
  ) +
  labs(
    title = "Dependents distribution for MTN customer base", 
    x = "dependents", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# Plotting a distribution by tenure
# Create the summary table
tenure_df <- Telecom_clean_df %>% 
  count(tenure)
tenure_df

# Plotting the chart
ggplot(tenure_df, aes(x = tenure, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#627909"
  ) +
  labs(
    title = "tenure distribution for MTN customer base", 
    x = "tenure", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# plotting distribution of phone service
# Create the summary table
phone_df <- Telecom_clean_df %>% 
  count(phoneservice)
phone_df

# Plotting the chart
ggplot(phone_df, aes(x = phoneservice, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#ff0009"
  ) +
  labs(
    title = "phone service distribution for MTN customer base", 
    x = "Phone Service", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
# Plotting a distribution of multiple lines
# Create the summary table
mult_df <- Telecom_clean_df %>% 
  count(multiplelines)
mult_df

# Plotting the chart
ggplot(mult_df, aes(x = multiplelines, y = n)) +
  geom_col(
    stat="identity", width=0.5, fill = "#ff8909"
  ) +
  labs(
    title = " Multiple Lines distribution for MTN customer base", 
    x = "Multiple Lines", 
    y = "Subscribers"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Summary of Findings

Based on the results of the analysis, the following conclusions were arrived at:

  There is no significant difference in churn rate between male and female subscribers. So this is not an area management needs to worry about.
  Majority of the customers are not senior citizens so this makes this dataset biased and hard to identify whether being a senior citizen affects churn rate.
  Not having a partner increases the likelihood of churning.
  Not having dependents increases the likelihood of churning.
  there's no significant difference between customers with multiple lines and those that do not.
  Customers on a month to month contract have a higher propensity to churn compared to those with long term contract
  
  
  
## Recommendations

In order to create an effective customer retention program, management should take the following measures:

  Focus more on meeting the needs of non-senior citizens.
  Focus more on having customers that have partners and/or dependents since these people are less likely to churn. Alternatively, management can come up with services specifically designed for customers without parters and/or dependents. This would require additional research.
  Focus more on getting customers to long term contract



